Adding Sensitivity to 21\,cm Inteferometric Probes of Reionization by Optimizing Choice of Baselines

### 2017-05-28 Rotation paper/draft_working.tex
save

The epoch of reionization represents the last key
stage of our Universe's early evolution. Study of this event stands at
the intersection of cosmology and astrophysics. Understanding this
event not only serves as a scientific goal
of its own, but also as a gateway to crucial information
regarding fundamental physics of inflation, neutrino mass and phenomenology
of the first stars and galaxies (e.g. \citealt{LiuOpticalDepth, Liu2016b, Mao2008, DEw21cm, Bull2015, Oyama20131186} etc.).

Arguably the most promising observational probe of the epoch of reionization
comes from measurement of the ``spin-flip'' transition of neutral
hydrogen of characteristic wavelength 21\,cm \citep{Furlanetto2006181,PritchardLoeb}.
Other observational studies of reionization include for example Gunn-Peterson measurements of quasi-stellar objects \citep{Fan2006}. GP tests rely on Ly$\alpha$ absorption, and thus easily saturate with small degree of ionization and can only probe the end stage of reionization. The 21\,cm line can in comparison directly probe the
abundance and distribution of neutral hydrogen, and thus would potentially
shed light on all stages of reionization. Radio interferometric efforts
to measure the 21\,cm power spectrum has been a top priority in recent years of astronomy.
Current-generation instruments include the Precision Array for Probing
the Epoch of Reionization (PAPER) \citep{Ali2015,paper32}, Murchison
Widefield Array (MWA) \citep{Bowman2013, Tingay2013}, Low Frequency Array (LOFAR) \citep{LOFAR}. Next-generation instruments include the Hydrogen Epoch of Reionization
Array (HERA) (e.g. \citealt{HERA,HERAconfiguration,HERABEAM1,HERADISH2})  under construction, 
and the Square Kilometer Array Low (SKA-low) (e.g. \citealt{SKA1}), which is currently in planning.

These modern low-frequency radio interferometers aimed at measuring the 21\,cm signal bears designs dramatically different from traditional instruments built for imaging localized bright sources.  With the goal of detecting a cosmological signal 5 orders of magnitude fainter than foreground contaminations, current efforts are sensitivity starved. Maximum-redundancy arrays are thus designed to repeatedly measure the same Fourier signal to increase sensitivity. Ranging upwards from 100 elements, these arrays are typically of the drift-scan type to limit cost.

Analysis pipelines of such maximum-redundancy arrays typically fall into two categories. In the first, more traditional technique, image is formed in Fourier domain through rotation synthesis, then foreground is either removed or avoided to construct a power spectrum. The second, lately popular technique works directly with visibilities from baselines and cross multiply them to form the power spectrum. Due to its more natural focus on the instrumental output of visibilities, this technique is shown to fair better in dealing with systematics such as RFIs and instrumental cross-talks, and thus tend to provide good limits on the power spectrum. \cite{Ali2015} provides the newest upper limit to the power spectrum
measurements with the 64-element version of PAPER (henceforth referred to as PAPER-64). However, one disadvantage of existing visibility-based pipelines is the lack of use of partial redundancy. Baselines of different length and orientation still contain partially redundant information. While an imaging based power-spectrum pipeline naturally includes all redundancy information, visibility-based pipelines until now only cross multiply fully redundant baselines, in other words baselines of the same length and orientation.

The use of partially redundant baselines is in principle not a fundamental limitation of visibility-based pipelines. In fact, most instrumental reports to date do include partial redundancy. Recently, \cite{wterm} proposed a visibility-based approach to extract power spectrum from partially redundant baselines in tracking measurements. Our work parallels this effort by focusing on arrays of the drift-scan type that do not have tracking capabilities, such as PAPER, HERA and SKA.
The Earth's rotation causes the baselines in a drift-scan array to pick up different modes of the sky with time. Rotation-synthesis makes use of the rotation-induced $uv$ coverage map to form image. In a visibility based pipeline, we can extract the same rotation-induced redundancy. In this point of view, baselines that are slightly different in length and orientation
``rotate into'' each other at a time delay. We can thus cross-multiply time-shifted visibilities, with the proper weighting, to form power spectrum.  Due to the large number of elements of modern arrays (upward of 100), the task of cross-multiplying every baseline against every other, scaling as number of array element to the fourth power, can be computationally formidable, and many pairs of baselines provide only negligible redundancy information. Our contribution is thus twofold. First we introduce a formalism to include pairs of partially-redundant baselines in a visibility based power spectrum pipeline. Secondly we show how to use the formalism to automatically pre-select baseline pairs and time offsets, making the problem computationally efficient. More precisely, our formalism allows one to simultaneously a) identify the baselines that give good
redundancy, b) find the time offset that corresponds to maximal redundancy for a given pair of baselines, and
c) quantify the sensitivity associated with cross multiplying
such a pair of partially redundant baselines, which in turns is used as weight to combine measurements in a power spectrum pipeline.

>>>>

The epoch of reionization represents the last key
stage of our Universe's early evolution. Study of this event stands at
the intersection of cosmology and astrophysics. Understanding this
event not only serves as a scientific goal
of its own, but also as a gateway to crucial information
regarding fundamental physics of inflation, neutrino mass and phenomenology
of the first stars and galaxies (e.g. \citealt{LiuOpticalDepth, Liu2016b, Mao2008, DEw21cm, Bull2015, Oyama20131186}).

Arguably the most promising observational probe of the epoch of reionization
comes from measurement of the ``spin-flip'' transition of neutral
hydrogen of characteristic wavelength 21\,cm \citep{Furlanetto2006181,PritchardLoeb}.
Other observational studies of reionization include for example Gunn-Peterson measurements of quasi-stellar objects \citep{Fan2006}. GP tests rely on Ly$\alpha$ absorption, and thus easily saturate with small degree of ionization and can only probe the end stage of reionization. [other probes include cmb optical depth, cmb ksz, lyman alpha emitters (redshifted as in bubble to no longer be absorbed, big bubbles corresponding to clustered sources), GP, UV luminosity of high redshift galaxy] The 21\,cm brightness temperature distribution is a direct tracer of neutral hydrogen through the epoch of reionization, and thus measurements of the 21\,cm line would potentially
shed light on all stages of reionization. Radio interferometric efforts
to measure the 21\,cm power spectrum [spatial pspec of brightness temp fluctuations] has been a top priority in recent years of astronomy.
Current-generation instruments include the Precision Array for Probing
the Epoch of Reionization (PAPER) \citep{Ali2015,paper32}, Murchison
Widefield Array (MWA) \citep{Bowman2013, Tingay2013}, Low Frequency Array (LOFAR) \citep{LOFAR}. Next-generation instruments include the Hydrogen Epoch of Reionization
Array (HERA) (e.g. \citealt{HERA,HERAconfiguration,HERABEAM1,HERADISH2})  under construction, 
and the Square Kilometer Array Low (SKA-low) (e.g. \citealt{SKA1}), which is currently in planning.

The highly redshifted 21\,cm signal is faint and diffuse, in contrast to the localized bright sources that have been of interest in many traditional applications of radio astronomy. Thus modern low-frequency radio interferometers aimed at measuring the 21\,cm signal bears designs different from traditional instruments.  Current experiments are sensitivity-starved, with the sensitivity requirement further increased considering the foreground contaminations 5 orders of manitude brighter. With the goal of detecting a cosmological signal 5 orders of magnitude fainter than foreground contaminations, current efforts are sensitivity starved. The arrays are thus designed with multiple copies of the same baselines to repeatedly measure the same Fourier signal to increase sensitivity. To satisfy the sensitivity need, modern arrays are large. Ranging upwards from 100 elements, these arrays are typically of the drift-scan type to limit cost.

Analysis pipelines for these new arrays typically fall into two categories. In the first, more traditional technique, images are formed in Fourier domain through rotation synthesis, followed by a foreground mitigation step to construct a power spectrum. An alternative technique works directly with visibilities from baselines and cross multiplies them to form the power spectrum. Tracking closely the instrumental output of visibilities, this technique allows one to deal with systematics such as Radio-Frequency-Interference(RFI) and instrumental cross-talks in a more native context, and thus tend to provide good limits on the power spectrum. An example of the visibility-based pipeline was presented in \cite{Ali2015}, which provides the newest upper limit to the power spectrum
measurements with the 64-element version of PAPER (henceforth as PAPER-64). However, one disadvantage of existing visibility-based pipelines is the lack of use of partial redundancy. Baselines of different length and orientation still contain partially redundant information. While an imaging based power-spectrum pipeline naturally includes all redundancy information, visibility-based pipelines until now only cross multiply fully redundant baselines, in other words baselines of the same length and orientation.

The use of partially redundant baselines is in principle not a fundamental limitation of visibility-based pipelines. In fact, most instrumental reports to date do include partial redundancy. Recently, \cite{wterm} proposed a visibility-based approach to extract power spectrum from partially redundant baselines in tracking measurements. Arrays capable of tracking include MWA and LOFAR. Our work parallels this effort by focusing on drift-scan arrays that do not have tracking capabilities, such as PAPER, HERA and potentially SKA.
The Earth's rotation causes the baselines in a drift-scan array to pick up different modes of the sky with time. Rotation-synthesis makes use of the rotation-induced $uv$ coverage map to form image. In a visibility based pipeline, we can extract the same rotation-induced redundancy. In this point of view, baselines that are slightly different in length and orientation
``rotate into'' each other at a time delay. We can thus cross-multiply time-shifted visibilities, with the proper weighting, to form power spectra.  Due to the large number of elements of modern arrays (upward of 100), the task of cross-multiplying every baseline against every other, scaling as number of array elements to the fourth power, can be computationally formidable, and many pairs of baselines provide only negligible redundancy information. Our contribution is thus twofold. First we introduce a formalism to include pairs of partially-redundant baselines in a visibility based power spectrum pipeline. Secondly we show how to use the formalism to automatically pre-select baseline pairs and time offsets, making the problem computationally efficient. More precisely, our formalism allows one to simultaneously identify the baselines that give good
redundancy, find the time offset that corresponds to maximal redundancy for a given pair of baselines, and quantify the sensitivity associated with cross multiplying
such a pair of partially redundant baselines, which in turns is used as weight to combine measurements in a power spectrum pipeline.
